### 2022-04-06 manuscript.tex
figurebold [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/cfeaf4b301d421f38362c16fb408e8a598ee88be.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/cfeaf4b301d421f38362c16fb408e8a598ee88be/checks
https://github.com/scherzerthesis/thesis/actions/runs/2104742536

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/3bd48ef9b7e8de9373449f399d612df876396338/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/3bd48ef9b7e8de9373449f399d612df876396338}{scherzerthesis/thesis@3bd48ef}
on April 6, 2022.

@@

Adenoviral vectors and lentiviral vectors introduced in the lungs of GEMMs offer temporal control and can be titrated to achieve an optimal level of tumor numerosity.
However, only lentiviral vectors can be used to barcode tumors because they integrate into the genome while adenoviruses do not(DuPage et al., 2009; Rogers et al., 2017).
With CRISPR/CAS9 gene editing, lentiviruses have been used to analyze many genetic interactions of oncogenic \emph{Kras} driven lung adenocarcinoma(Rogers et al., 2017; Winters et al., 2017).
To enable simultaneous activation of BRAF\textsuperscript{V600E} and loss of tumor suppressor genes, we bred \emph{Braf\textsuperscript{CAT}}; \emph{H11b\textsuperscript{LSL-CAS9/+}} mice.
To determine the size of BRAF \textsuperscript{V600E} tumors and the contribution of 11 tumor suppressor genes (\emph{Trp53, Rb1, Cdkn2a, Apc, Rbm10, Stk11, Setd2, Arid1a, Atm, Keap1, Smad4)} on BRAF \textsuperscript{V600E} lung tumorigenesis, we obtained tittered barcoded-lentiviral plasmids that encode Cre-recombinase as well as a sgRNA that inactivates one of eleven tumor suppressors common in lung adenocarcinoma.
All genes tested are putative tumor suppressor genes commonly lost in human lung cancer patients. Each lentiviral solution contained \textasciitilde1000 individually barcoded lentiviruses.
We infected both \emph{LSL-Cas9} and \emph{CAS9} deficient mice in order to determine the contribution of genetic manipulation on tumor size.
Mice meant to be analyzed after eight weeks received a viral titer of 1x10\^{}5, while mice meant to be analyzed after 16 weeks received a viral titer of 1x10\^{}4. After euthanasia, lungs were collected for genomic DNA extraction and barcode amplification, followed by the preparation of sequencing libraries.
Before genomic extraction and library preparation, a known quantity of known-barcoded cells was added to benchmark sequencing reads to a known cell number(figure).

We determined the mean size of a BRAF\textsuperscript{V600E} driven tumor to be 21,248 thousand cells (figure).
Interestingly, there is a widespread distribution of tumor sizes, with the most giant BRAF\textsuperscript{V600E} tumor to be \textasciitilde300,000 cells, along with many small tumors (figure).

Gross, histologic images of experimental lungs show an increase in tumor burden with both time (Figure xxx) and CAS9 status (figure xxx).
Therefore, based on macroscopic inspection, at least one tumor suppressor appeared to cooperate with BRAF \textsuperscript{V600E} to increase tumor burden.
In order to determine the fitness landscape of any one particular tumor suppressor, we calculated the relative number of tumor cells in a given tumor using the Tuba-Seq genomics pipeline.
After eight weeks post tumor initiation, there is no statistically significant contribution of tumor suppressor loss (Figure).
SETD2 loss was trending towards statistical significance.
However, 16 weeks post-initiation revealed that both SETD2 loss and APC loss cooperated with BRAF \textsuperscript{V600E} to form larger tumors (Figure x).
APC loss was a suspected hit as we have previously shown that activation of Wnt-beta-catenin signaling promoted BRAF\textsuperscript{V600E} tumorigenesis.
As APC loss is predicted to increase Beta-catenin signaling activity, this result was unsurprising to us.
However, loss of SETD2 was a novel finding in the context of BRAF \textsuperscript{V600E} lung tumorigenesis.

Given our findings of SETD2 loss increased tumor burden concurrent with BRAF\textsuperscript{V600E}, we next validated SETD2 in a new cohort of mice with single-guide RNAs targeting SETD2.
Indeed, loss of SETD2 increased tumor burden from 5\% to over 30\% in some animals.
Since SETD2 antibodies are unreliable in both western blots and immunohistochemistry, we used H3K36me3 as a surrogate marker for SETD2 activity by immunohistochemistry.
Whereas 90\% of nuclei in BRAF \textsuperscript{V600E} only tumors were H3K36me3, more than 60\% of tumors of mice infected with SETD2 CRISPR virus were negative for H3K36me3.
Furthermore, when classified by H3K36me3-positive or -negative, there is even more substantial evidence that SETD2 loss leads to larger tumor formation.

@@

Expression of BRAF\textsuperscript{V600E} in the distal epithelium of the mouse lung elicits clonal tumorigenic outgrowths of alveolar type 2 (AT2) pneumocytes (Dankort et al., 2007). Prior work has shown that \emph{Trp53\textsuperscript{fl/fl}} cooperates with BRAF(V600E) to form malignant lung tumors when Cre recombinase is activated (Dankort et al., 2007; Shai et al., 2015). Therefore, we compared the tumor burden of \emph{Braf\textsuperscript{Cat/+}} mice with either p53 loss (\emph{Trp53\textsuperscript{fl/fl}}, p53\textsuperscript{R172H} (\emph{Trp53\textsuperscript{R172H/R172H}}), or p53\textsuperscript{R245W} (\emph{Trp53\textsuperscript{R245W/R245W}}) after infection with 106 pfu of adenovirus expressing surfactant protein C specific Cre (ad5-Spc-Cre) (Figure \ref{fig:3.1}). As expected, mice developed lung cancer, consistent with previous findings (Figure \ref{fig:3.2} B-C). However, there was a striking, highly reproducible increase in the tumor size of tumors harboring \emph{Trp53\textsuperscript{R172H/R172H}} and to a lesser extent \emph{Trp53\textsuperscript{R245W/R245W}}, compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure \ref{fig:3.2} A,B). These data led us to hypothesize that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} expression alters tumor progression uniquely from p53 deletion, and these differences may be potentially evident of gain-of-function effects imparted by mutant-53. We next looked at early time points to determine if p53 mutation affects the growth and proliferation of early stage lung tumors. However, we did not find any differences in tumor size between the three different p53 alleles(Figure (\textbf{fig3.3?}) A, B).

Since lung cancer patients with mutations in \emph{TP53} often start only have one mutated \emph{p53} allele unless the loss of heterozygosity (LOH) occurs. Therefore, we generated mice heterozygous for both \emph{Braf\textsuperscript{CAT}} and either \emph{Trp53\textsuperscript{fl}}, \emph{Trp53\textsuperscript{R172H}}, or \emph{Trp53\^{} R245W\^{}} and observed their survival over time (Figure \ref{fig:3.4} A). We noted that \emph{Trp53\textsuperscript{R172H/+}} mice had diminished median overall survival compared to \emph{Trp53\textsuperscript{fl/+}} mice (183 days and 216 days), with \emph{Trp53\textsuperscript{R245W/+}} mice falling in between the two groups (median survival 212 days) (Figure \ref{fig:3.4} A).

We observed a distinct separation of survival between mice harboring lung tumors that are either \emph{Trp53\textsuperscript{fl/+}}, \emph{Trp53\textsuperscript{R172H/+}}, and \emph{Trp53\textsuperscript{R245W/+}}. This separation is diminished when altered p53 alleles are homozygous (Figure \ref{fig:3.4} B). \emph{Trp53\textsuperscript{R172H/R172H}} mice have a median survival of 120 days, while \emph{Trp53\textsuperscript{fl/fl}} mice have a median survival of 136 days. Importantly, this observed difference in survival is statistically significant (Log-rank test p\textless{} 0.001) and is likely to reflect the worse diseased, characterized by enhanced tumor burden and size.

\textbf{Mutant-p53 Accumulation is Associated with Dedifferentiation and p19 Induction}
Increased malignancy in lung adenocarcinoma is often associated with altered states of tumor cell identity and dedifferentiation (Marjanovic et al., 2020; van Veen et al., 2019). Since LUADs often arise from AT2 cells, we surveyed the expression of AT2 cell identifiers, such as NKX2-1, a master regulator of AT2 transcriptional programs, and SPC, a marker of differentiated AT2 cells (Marjanovic et al., 2020). Quantification of over 50 tumors from \emph{Braf\textsuperscript{V600E/+}} mice with either \emph{Trp53\^{} flfl\^{}}, \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} revealed that \emph{Trp53\textsuperscript{R172H/R1722H}} mice had significantly more tumors that stained negative for NKX2-1 and SPC (Figure \ref{fig:3.5} A, B). This finding strongly indicates that mutant-p53 tumors are dedifferentiated. Since HMGA2 expression is associated with LUAD dedifferentiation and worse patient prognosis, we performed IHC for HMGA2(Di Cello et al., 2008; Gao et al., 2017; Winslow et al., 2011). We noted an increase in HMGA2-positive tumors in \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} tumors compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure \ref{fig:3.5} A, B). Thus, this data reveals that Trp53R172H/R172H mice exhibit a higher tumor burden and worse overall survival than \emph{Trp53\textsuperscript{fl/fl}}. However, the tumors developed in these mice are more likely to exhibit malignant progression and dedifferentiation.

One striking feature of lung tumors in mice, either heterozygous or homozygous for \emph{Trp53 \textsuperscript{R172H}} or \emph{Trp53\textsuperscript{R245W}} is how few tumors and tumor cells have detectable levels of mutant-p53 protein as measured by Immunohistochemistry. One critical aspect of normal p53 function is self-regulation through transcriptional activation of its negative regulator MDM2 (Wu et al., 1993). Thus, we predicted that homozygous p53 mutant cells would exhibit increased mutant-p53 protein. In our models, we observed that mutant p53 protein accumulation is restricted to malignant cells exhibited by a lack of papillary architecture and nuclear atypia (Figure \ref{fig:3.6} A). Indeed, p53 immunostaining is correlated with increased phosphorylated-ERK1/2 detection (Figure \ref{fig:3.6} B).
Moreover, immunofluorescence of p53 and its positive regulator, p19/ARF, reveal that cells with high p19/ARF are also high in p53 accumulation (Figure \ref{fig:3.6} C). This result suggests that even in homozygous mutant\_Trp53\_ mice, there is still a barrier to mutant-p53 accumulation. Our data suggest that elevated levels of p19/ARF are necessary for mutant-p53 accumulation, even with homozygous p53 mutations. Collectively, these data indicate that \emph{Trp53\textsuperscript{R172H}} drives more aggressive lung adenocarcinoma formation in BRAF(V600E) driven mouse models of lung cancer.

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} cell lines are refractory to BRAF/MEK inhibition}
Given that previous research has determined that p53 mutations can confer resistance to both chemotherapies and pathway-targeted therapies (Huang et al., 2011; Rusch et al., 1995), we next questioned whether our p53 alterations would be sufficient to drive therapeutic resistance in BRAF\textsuperscript{V600E}-driven genetically engineered mice(Planchard et al., 2016). Interstingly, \emph{Trp53\textsuperscript{R172H/R172H}} mice retain a more considerable tumor burden after a week four treatment of MAPK blockade \emph{in vivo} (Figure \ref{fig:3.7} A).

To further investigate the effects of p53-mutants on \emph{BRAF\textsuperscript{V600E}} driven lung cancers,we derived cell lines from \emph{Braf\textsuperscript{CAT/+}} murine tumors expressing either \emph{Trp53\textsuperscript{flfl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}) 14 weeks after infection with ad5-Spc-Cre. Cell lines expressing either \emph{Trp53\textsuperscript{R172H/R172H}} or \emph{Trp53\textsuperscript{R245W/R245W}} were resistant to single-agent dabrafenib (Figure \ref{fig:3.7} B). Of note, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines express increased ERK1/2 activity, as evidenced by increased phosphorylated ERK1/2, both with and without MAPK-blockade (Figure \ref{fig:3.7} C). Furthermore, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} were more resistant to the combination of dabrafenib plus trametinib, than \emph{Trp53\textsuperscript{fl/fl}} cell lines(Figure \ref{fig:3.8} A-C). Not only are resistant cells still proliferative in the face of MAPK-blockade, they are resistant to cell death imparted by MAPK-inhibtion{[}Figure \ref{fig:3.8} C{]}.

Next, we reasoned if mutant-p53 expression is the cause differential cancer phenotypes such as increase proliferation and drug resistance, then eliminating mutant-p53 expression would have a growth inhibitory effect and sensitize cells to MAPK-pathway inhibition. We cloned a short-hairpin RNA plasmid that induces GFP expression and knockdown of mutant-p53 with supplementation of doxycycline(Figure \ref{fig:3.9} A). Despite sucessful p53 knockdown, observed no change in \emph{in vitro} proliferation or \emph{in vivo} tumor outgrowth(Figure \ref{fig:3.9} B and C). Therefore, mutant-p53 cells are not addicted to mutant-p53 expression.

To gain relevance to human lung cancer, we infected BRAF(V600E) mutant HCC364 cells with lentivirus expressing human cancer hotspot mutant p53\textsuperscript{R175H}. p53\textsuperscript{R175H} expressing cells were more proliferative and refractory than parental HCC364 cells (Figure \ref{fig:3.10} A and B). Collectively, these \emph{in vitro} and \emph{in vivo} data suggest that missense mutant p53\textsuperscript{R175H} and its murine homolog p53\textsuperscript{R172H}, as well as p53\textsuperscript{R245W} can support cancer cell viability in the face of BRAF+MEK blockade that p53-null mutations cannot achieve.

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} alter the transcriptome of BRAF(V600E) lung tumor cell lines}
Due to our observations that mutant p53\textsuperscript{R172H} drives more aggressive cancer phenotypes such as increased growth, dedifferentiation, and resistance to FDA-approved therapeutics, we reasoned that this would be reflected in a change in the transcriptional landscape. This is due in part to the transcriptional trans-activation potential by mutant p53. RNA sequencing of cell lines generated from p53-mutant mouse lung tumors does reveal transcriptomic differences between both p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cells compared to p53-null cell lines (Figure \ref{fig:3.11} 4 A and B). Interestingly, genes involved in cell cycle regulation and MYC pathway are upregulated in both p53\textsuperscript{R172H} and p53\textsuperscript{R2245W} cell lines (Figure 4 \ref{fig:3.11} B). These results suggest that mutant-p53 affects the transcription of genes involved in pro-growth signaling pathways. We also observed transcriptional differences between \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines(Figure \ref{fig:3.12} A and B). Thus, we hypothesize that \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} have similarities in their function, but they also have some clear differences.

\textbf{PDGFR expression is increased in mutant-p53 cells}
Our \emph{in vitro} RNA analysis revealed an upregulation of the platelet-derived growth factor recepter(PDGFR) in \emph{Trp53\textsuperscript{R172H/R172H}} cell lines(Figure \ref{fig:3.13} A). Since PDGFR is a known regulator of the MAPK pathway, we sought to validate its expression and determine its important in \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines. Indeed PDGFRB is more highly expressed and can be activated by recombinant PDGF stimulation(Figure \ref{fig:3.14} B). Although \emph{Trp53\textsuperscript{fl/fl}} cells do not have any detectable levels of PDGFR or Phospohorylated-PDFRB, they still upregulated MAPK-pathway activation in response to ligand stimulation(Figure \ref{fig:3.14} B). Furthermore, expression of PDGFRB is evident in malignant \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} tumors, but not \emph{Trp53\textsuperscript{fl/fl}} tumors(Figure \ref{fig:3.14} C). We next sought to determine the sensitivity to PDGFR-signaling blockade with Imatinib, a known PDGFRB inhibitor. Neither PDGFR stimulation or PDGFR inhibition had any effect on the proliferation of mutant-p53 cell lines(Figure \ref{fig:3.15} A an B). Therefore, PDGFR is a validated differentially expressed gene in \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines, but is not important for its growth.

@@

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/cfeaf4b301d421f38362c16fb408e8a598ee88be/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/cfeaf4b301d421f38362c16fb408e8a598ee88be}{scherzerthesis/thesis@cfeaf4b}
on April 6, 2022.

@@

Adenoviral vectors and lentiviral vectors introduced in the lungs of GEMMs offer temporal control and can be titrated to achieve an optimal level of tumor numerosity.
However, only lentiviral vectors can be used to barcode tumors because they integrate into the genome while adenoviruses do not(DuPage et al., 2009; Rogers et al., 2017).
With CRISPR/CAS9 gene editing, lentiviruses have been used to analyze many genetic interactions of oncogenic \emph{Kras} driven lung adenocarcinoma(Rogers et al., 2017; Winters et al., 2017).
To enable simultaneous activation of BRAF\textsuperscript{V600E} and loss of tumor suppressor genes, we bred \emph{Braf\textsuperscript{CAT}}; \emph{H11b\textsuperscript{LSL-CAS9/+}} mice.
To determine the size of BRAF \textsuperscript{V600E} tumors and the contribution of 11 tumor suppressor genes (\emph{Trp53, Rb1, Cdkn2a, Apc, Rbm10, Stk11, Setd2, Arid1a, Atm, Keap1, Smad4)} on BRAF \textsuperscript{V600E} lung tumorigenesis, we obtained tittered barcoded-lentiviral plasmids that encode Cre-recombinase as well as a sgRNA that inactivates one of eleven tumor suppressors common in lung adenocarcinoma(Figure 2.1).
All genes tested are putative tumor suppressor genes commonly lost in human lung cancer patients. Each lentiviral solution contained \textasciitilde1000 individually barcoded lentiviruses.
We infected both \emph{LSL-Cas9} and \emph{CAS9} deficient mice in order to determine the contribution of genetic manipulation on tumor size (Figure 2.1).
Mice meant to be analyzed after eight weeks received a viral titer of 1x10\^{}5, while mice meant to be analyzed after 16 weeks received a viral titer of 1x10\^{}4.
Gross, histologic images of experimental lungs show an increase in tumor burden with both time (Figure xxx) and CAS9 status (Figure 2.1).
Therefore, based on macroscopic inspection, at least one tumor suppressor appeared to cooperate with BRAF \textsuperscript{V600E} to increase tumor burden.
After euthanasia, lungs were collected for genomic DNA extraction and barcode amplification, followed by the preparation of sequencing libraries.
Before genomic extraction and library preparation, a known quantity of known-barcoded cells was added to benchmark sequencing reads to a known cell number.

We determined the mean size of a BRAF\textsuperscript{V600E} driven tumor to be 21,248 thousand cells (Figure 2.2).
Interestingly, there is a widespread distribution of tumor sizes, with the most giant BRAF\textsuperscript{V600E} tumor to be \textasciitilde300,000 cells, along with many small tumors (Figure).

In order to determine the fitness landscape of any one particular tumor suppressor, we calculated the relative number of tumor cells in a given tumor using the Tuba-Seq genomics pipeline. In CAS9 deficient mice, there was no statiscically significant cooperation with any single-guide RNAs (Figure 2.3A).
After eight weeks post tumor initiation, APC loss provided a growth advantage(Figure 2.3 B).
However, 16 weeks post-initiation revealed that both SETD2 loss and APC loss cooperated with BRAF \textsuperscript{V600E} to form larger tumors (Figure 2.3 C).
APC loss was a suspected hit as we have previously shown that activation of Wnt-beta-catenin signaling promoted BRAF\textsuperscript{V600E} tumorigenesis.
As APC loss is predicted to increase Beta-catenin signaling activity, this result was unsurprising to us.
However, loss of SETD2 was a novel finding in the context of BRAF \textsuperscript{V600E} lung tumorigenesis.

Given our findings of SETD2 loss increased tumor burden concurrent with BRAF\textsuperscript{V600E}, we next validated SETD2 in a new cohort of mice with single-guide RNAs targeting SETD2 (Figure 2.4).
Indeed, loss of SETD2 increased tumor size based on gross histological analysis(Figure 2.4A). Addiionally, tumor size increased from in reponse to CRISPR/Cas9-mediated gene editing of \emph{Setd2}(Figure 2.4B).
Since SETD2 antibodies are unreliable in both western blots and immunohistochemistry, we used H3K36me3 as a surrogate marker for SETD2 activity by immunohistochemistry(Figure 2.5A).
Whereas 90\% of nuclei in BRAF \textsuperscript{V600E} only tumors were H3K36me3, more than 60\% of tumors of mice infected with SETD2 CRISPR virus were negative for H3K36me3(Figure 2.5B).
Furthermore, when classified by H3K36me3-positive or -negative, there is even more substantial evidence that SETD2 loss leads to larger tumor formation(Figure 2.5C). Specifically, tumors that have lost H3H36me\textsuperscript{3}, were 5-times larger than tumors that have retained H3K36me\textsuperscript{3}. Therefore, we conclude that \emph{Setd2} loss drives growth of BRAF\textsuperscript{V600E}-driven lung tumors.

Clearly, there is a strong selective advantage given to tumors that have lost \emph{Setd2} function. We next wondered if human lung cancer patients have altered \emph{SETD2} levels. Indeed, Low \emph{SETD2} levels correlate with poor overall survival and \emph{SETD2} is expressed less in lung tumors compared to normal lung tissue(Figure 2.6A,B). To further demonstrate the relationship between BRAF\textsuperscript{V600E} and \emph{SETD2} loss, we mined patient data avaiable from Project Genie(2017). Interestingly, we found an association between \emph{SETD2} genomic alterations and \emph{\textsuperscript{BRAF}V600\^{}} mutational activation(Figure 2.6C). Specifically, of the 220 \emph{BRAF\textsuperscript{V600}} mutations (include V600E and V600K), 40\% had concurrent alterations in \emph{SETD2}. These observations indicate \emph{SETD2} loss is a common feature of \emph{BRAF\textsuperscript{V600}}-driven human lung adenocarcinomas and predicts poor overall survival in non-small cell lung cancer.

@@

Expression of BRAF\textsuperscript{V600E} in the distal epithelium of the mouse lung elicits clonal tumorigenic outgrowths of alveolar type 2 (AT2) pneumocytes (Dankort et al., 2007). Prior work has shown that \emph{Trp53\textsuperscript{fl/fl}} cooperates with BRAF(V600E) to form malignant lung tumors when Cre recombinase is activated (Dankort et al., 2007; Shai et al., 2015). Therefore, we compared the tumor burden of \emph{Braf\textsuperscript{Cat/+}} mice with either p53 loss (\emph{Trp53\textsuperscript{fl/fl}}, p53\textsuperscript{R172H} (\emph{Trp53\textsuperscript{R172H/R172H}}), or p53\textsuperscript{R245W} (\emph{Trp53\textsuperscript{R245W/R245W}}) after infection with 106 pfu of adenovirus expressing surfactant protein C specific Cre (ad5-Spc-Cre) (Figure 3.1). As expected, mice developed lung cancer, consistent with previous findings (Figure 3.2B-C). However, there was a striking, highly reproducible increase in the tumor size of tumors harboring \emph{Trp53\textsuperscript{R172H/R172H}} and to a lesser extent \emph{Trp53\textsuperscript{R245W/R245W}}, compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure 3.2A and B). These data led us to hypothesize that p53\textsuperscript{R172H} and p53\textsuperscript{R245W} expression alters tumor progression uniquely from p53 deletion, and these differences may be potentially evident of gain-of-function effects imparted by mutant-53. We next looked at early time points to determine if p53 mutation affects the growth and proliferation of early stage lung tumors. However, we did not find any differences in tumor size between the three different p53 alleles(Figure 3.3A and B).

Since lung cancer patients with mutations in \emph{TP53} often start only have one mutated \emph{p53} allele unless the loss of heterozygosity (LOH) occurs. Therefore, we generated mice heterozygous for both \emph{Braf\textsuperscript{CAT}} and either \emph{Trp53\textsuperscript{fl}}, \emph{Trp53\textsuperscript{R172H}}, or \emph{Trp53\^{} R245W\^{}} and observed their survival over time (Figure 3.4A). We noted that \emph{Trp53\textsuperscript{R172H/+}} mice had diminished median overall survival compared to \emph{Trp53\textsuperscript{fl/+}} mice (183 days and 216 days), with \emph{Trp53\textsuperscript{R245W/+}} mice falling in between the two groups (median survival 212 days) (Figure 3.4A).

We observed a distinct separation of survival between mice harboring lung tumors that are either \emph{Trp53\textsuperscript{fl/+}}, \emph{Trp53\textsuperscript{R172H/+}}, and \emph{Trp53\textsuperscript{R245W/+}}. This separation is diminished when altered p53 alleles are homozygous (Figure 3.4B). \emph{Trp53\textsuperscript{R172H/R172H}} mice have a median survival of 120 days, while \emph{Trp53\textsuperscript{fl/fl}} mice have a median survival of 136 days. Importantly, this observed difference in survival is statistically significant (Log-rank test p\textless{} 0.001) and is likely to reflect the worse diseased, characterized by enhanced tumor burden and size.

\textbf{Mutant-p53 Accumulation is Associated with Dedifferentiation and p19 Induction}
Increased malignancy in lung adenocarcinoma is often associated with altered states of tumor cell identity and dedifferentiation (Marjanovic et al., 2020; van Veen et al., 2019). Since LUADs often arise from AT2 cells, we surveyed the expression of AT2 cell identifiers, such as NKX2-1, a master regulator of AT2 transcriptional programs, and SPC, a marker of differentiated AT2 cells (Marjanovic et al., 2020). Quantification of over 50 tumors from \emph{Braf\textsuperscript{V600E/+}} mice with either \emph{Trp53\^{} flfl\^{}}, \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} revealed that \emph{Trp53\textsuperscript{R172H/R1722H}} mice had significantly more tumors that stained negative for NKX2-1 and SPC (Figure 3.5 A and B). This finding strongly indicates that mutant-p53 tumors are dedifferentiated. Since HMGA2 expression is associated with LUAD dedifferentiation and worse patient prognosis, we performed IHC for HMGA2(Di Cello et al., 2008; Gao et al., 2017; Winslow et al., 2011). We noted an increase in HMGA2-positive tumors in \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} tumors compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure 3.5A and B). Thus, this data reveals that Trp53R172H/R172H mice exhibit a higher tumor burden and worse overall survival than \emph{Trp53\textsuperscript{fl/fl}}. However, the tumors developed in these mice are more likely to exhibit malignant progression and dedifferentiation.

One striking feature of lung tumors in mice, either heterozygous or homozygous for \emph{Trp53 \textsuperscript{R172H}} or \emph{Trp53\textsuperscript{R245W}} is how few tumors and tumor cells have detectable levels of mutant-p53 protein as measured by Immunohistochemistry. One critical aspect of normal p53 function is self-regulation through transcriptional activation of its negative regulator MDM2 (Wu et al., 1993). Thus, we predicted that homozygous p53 mutant cells would exhibit increased mutant-p53 protein. In our models, we observed that mutant p53 protein accumulation is restricted to malignant cells exhibited by a lack of papillary architecture and nuclear atypia (Figure 3.6A). Indeed, p53 immunostaining is correlated with increased phosphorylated-ERK1/2 detection (Figure 3.6B).
Moreover, immunofluorescence of p53 and its positive regulator, p19/ARF, reveal that cells with high p19/ARF are also high in p53 accumulation (Figure 3.6C). This result suggests that even in homozygous mutant\_Trp53\_ mice, there is still a barrier to mutant-p53 accumulation. Our data suggest that elevated levels of p19/ARF are necessary for mutant-p53 accumulation, even with homozygous p53 mutations. Collectively, these data indicate that \emph{Trp53\textsuperscript{R172H}} drives more aggressive lung adenocarcinoma formation in BRAF(V600E) driven mouse models of lung cancer.

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} cell lines are refractory to BRAF/MEK inhibition}
Given that previous research has determined that p53 mutations can confer resistance to both chemotherapies and pathway-targeted therapies (Huang et al., 2011; Rusch et al., 1995), we next questioned whether our p53 alterations would be sufficient to drive therapeutic resistance in BRAF\textsuperscript{V600E}-driven genetically engineered mice(Planchard et al., 2016). Interstingly, \emph{Trp53\textsuperscript{R172H/R172H}} mice retain a more considerable tumor burden after a week four treatment of MAPK blockade \emph{in vivo} (Figure 3.7A).

To further investigate the effects of p53-mutants on \emph{BRAF\textsuperscript{V600E}} driven lung cancers,we derived cell lines from \emph{Braf\textsuperscript{CAT/+}} murine tumors expressing either \emph{Trp53\textsuperscript{flfl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}) 14 weeks after infection with ad5-Spc-Cre. Cell lines expressing either \emph{Trp53\textsuperscript{R172H/R172H}} or \emph{Trp53\textsuperscript{R245W/R245W}} were resistant to single-agent dabrafenib (Figure 3.7B). Of note, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines express increased ERK1/2 activity, as evidenced by increased phosphorylated ERK1/2, both with and without MAPK-blockade (Figure 3.7C). Furthermore, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} were more resistant to the combination of dabrafenib plus trametinib, than \emph{Trp53\textsuperscript{fl/fl}} cell lines(Figure 3.8A-C). Not only are resistant cells still proliferative in the face of MAPK-blockade, they are resistant to cell death imparted by MAPK-inhibtion(Figure 3.8C).

Next, we reasoned if mutant-p53 expression is the cause differential cancer phenotypes such as increase proliferation and drug resistance, then eliminating mutant-p53 expression would have a growth inhibitory effect and sensitize cells to MAPK-pathway inhibition. We cloned a short-hairpin RNA plasmid that induces GFP expression and knockdown of mutant-p53 with supplementation of doxycycline(Figure 3.9A). Despite sucessful p53 knockdown, observed no change in \emph{in vitro} proliferation or \emph{in vivo} tumor outgrowth(Figure 3.9B and C). Therefore, mutant-p53 cells are not addicted to mutant-p53 expression.

To gain relevance to human lung cancer, we infected BRAF(V600E) mutant HCC364 cells with lentivirus expressing human cancer hotspot mutant p53\textsuperscript{R175H}. p53\textsuperscript{R175H} expressing cells were more proliferative and refractory than parental HCC364 cells (Figure 3.10A and B). Collectively, these \emph{in vitro} and \emph{in vivo} data suggest that missense mutant p53\textsuperscript{R175H} and its murine homolog p53\textsuperscript{R172H}, as well as p53\textsuperscript{R245W} can support cancer cell viability in the face of BRAF+MEK blockade that p53-null mutations cannot achieve.

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} alter the transcriptome of BRAF(V600E) lung tumor cell lines}
Due to our observations that mutant p53\textsuperscript{R172H} drives more aggressive cancer phenotypes such as increased growth, dedifferentiation, and resistance to FDA-approved therapeutics, we reasoned that this would be reflected in a change in the transcriptional landscape. This is due in part to the transcriptional trans-activation potential by mutant p53. RNA sequencing of cell lines generated from p53-mutant mouse lung tumors does reveal transcriptomic differences between both p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cells compared to p53-null cell lines (Figure 3.11A and B). Interestingly, genes involved in cell cycle regulation and MYC pathway are upregulated in both p53\textsuperscript{R172H} and p53\textsuperscript{R2245W} cell lines (Figure 3.11B). These results suggest that mutant-p53 affects the transcription of genes involved in pro-growth signaling pathways. We also observed transcriptional differences between \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines(Figure (\textbf{3.12A?}) and B). Thus, we hypothesize that \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} have similarities in their function, but they also have some clear differences.

\textbf{PDGFR expression is increased in mutant-p53 cells}
Our \emph{in vitro} RNA analysis revealed an upregulation of the platelet-derived growth factor recepter(PDGFR) in \emph{Trp53\textsuperscript{R172H/R172H}} cell lines(Figure 3.13A). Since PDGFR is a known regulator of the MAPK pathway, we sought to validate its expression and determine its important in \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines. Indeed PDGFRB is more highly expressed and can be activated by recombinant PDGF stimulation(Figure 3.14B). Although \emph{Trp53\textsuperscript{fl/fl}} cells do not have any detectable levels of PDGFR or Phospohorylated-PDFRB, they still upregulated MAPK-pathway activation in response to ligand stimulation(Figure 3.14B). Furthermore, expression of PDGFRB is evident in malignant \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} tumors, but not \emph{Trp53\textsuperscript{fl/fl}} tumors(Figure 3.14C). We next sought to determine the sensitivity to PDGFR-signaling blockade with Imatinib, a known PDGFRB inhibitor. Neither PDGFR stimulation or PDGFR inhibition had any effect on the proliferation of mutant-p53 cell lines(Figure 3.15A and B). Therefore, PDGFR is a validated differentially expressed gene in \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines, but is not important for its growth.

@@

\leavevmode\vadjust pre{\hypertarget{ref-1CW3EbE5O}{}}%
(2017). AACR Project GENIE: Powering Precision Medicine through an International Consortium. Cancer Discov \emph{7}, 818--831. \url{https://doi.org/10.1158/2159-8290.cd-17-0151}.
